Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-02-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: correction to Continuity AI and updates to bar plot of simd performance

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    10      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
Working-set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. $n$ is the problem size -- the length of one side of the
  square domain.}
\label{TAB_kernel_details}
\end{table}

@@

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1~MB and 4.2~MB, respectively and therefore
both of these should fit comfortably within L3 cache. Despite this,
the performance achieved with these domain sizes falls short of the L3
bandwidth limit.  Although this kernel is one of several within the
time-stepping loop of NEMOLite2D, the working-set size of this loop
for the $128^2$ domain is just 2.4~MB which again is well within the
capacity of L3.

@@

In order to investigate why the performance of the Continuity kernel
is falling well short of the maximum predicted by the roofline model
we have experimented with the level of SIMD
vectorisation. Figure~\ref{FIG_kernel_perf} shows the performance of
the Continuity and u-Momentum kernels when compiled without
vectorisation (``-no-vec''), with SSE4 (Streaming SIMD Extensions)
vectorisation (``-axSSE4.2'') and with AVX (``-xHost''). SSE supports
vector lengths of 128 bits (two double-precision words) and AVX 512
bits (four double-precision words). For the 64, 128 and 256 problem
sizes, changing from no vectorisation to SSE increases the performance
of the Continuity kernel by up to roughly 60\%; some way short of the
ideal speed-up of 100\% that would be obtained with perfect
SSE vectorisation of a compute-bound kernel. Unsurprisingly then, moving
from SSE to AVX vectors does not give any performance improvement and
in fact, is slightly detrimental for most of the problem sizes.

\begin{figure}
  \centering
  \includegraphics[width=120mm]{nemolite2d_kernel_perf}
  \caption{Performance of the Continuity and u-Momentum kernels as a
    function of SIMD vectorisation and problem size. Results are for
    v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.}
  \label{FIG_kernel_perf}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=120mm]{nemolite2d_cont_perf}
  \caption{Performance of the Continuity kernel (bars -- right y-axis)
    and HWPC data (open circles -- left y-axis) as a function of
    problem size. Results are for v.16 of the Intel compiler on a Xeon
    E5-1620 v.2 CPU.}
  \label{FIG_cont_perf}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=120mm]{continuity_opt}
  \caption{Performance of the Continuity kernel with optimisations
    intended to aid SIMD vectorisation. Results are for v.16 of the
    Intel compiler on a Xeon E5-1620 v.2 CPU.}
  \label{FIG_cont_vec}
\end{figure}

>>>>

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    12      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
Working-set size (bytes)   & $56n^2$ &  $72n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. $n$ is the problem size -- the length of one side of the
  square domain.}
\label{TAB_kernel_details}
\end{table}

@@

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1.1~MB and 4.7~MB, respectively and therefore
both of these should fit comfortably within L3 cache. Despite this,
the performance achieved with these domain sizes falls short of the L3
bandwidth limit.  Although this kernel is one of several within the
time-stepping loop of NEMOLite2D, the working-set size of the latter
for the $128^2$ domain is just 2.4~MB which again is well within the
capacity of L3.

@@

In order to investigate why the performance of the Continuity kernel
is falling well short of the maximum predicted by the roofline model
we have experimented with the level of SIMD
vectorisation. Figure~\ref{FIG_cont_vec} shows the performance of the
Continuity kernel compiled with differing levels of SIMD vectorisation
and after code modifications intended to improve vectorisation.
Compiling without auto-vectorisation (``-no-vec''), the kernel
achieves a performance that is largely limited by the bandwidth to
main memory. Note that there is very little opportunity for temporal
cache re-use in the Continuity kernel -- only three of the twelve
distinct memory references are to previous rows of the array. The size
of this effect can be seen in the small drop in performance (black
bars) in going from the $256^2$ to the $512^2$ domain as at this point
the working-set size spills out of L3 cache.

Figure~\ref{FIG_kernel_perf} shows the performance of
the Continuity and u-Momentum kernels when compiled without
vectorisation (``-no-vec''), with SSE4 (Streaming SIMD Extensions)
vectorisation (``-axSSE4.2'') and with AVX (``-xHost''). SSE supports
vector lengths of 128 bits (two double-precision words) and AVX 512
bits (four double-precision words). For the 64, 128 and 256 problem
sizes, changing from no vectorisation to SSE increases the performance
of the Continuity kernel by up to roughly 60\%; some way short of the
ideal speed-up of 100\% that would be obtained with perfect
SSE vectorisation of a compute-bound kernel. Unsurprisingly then, moving
from SSE to AVX vectors does not give any performance improvement and
in fact, is slightly detrimental for most of the problem sizes.

\begin{figure}
  \centering
  \includegraphics[width=120mm]{nemolite2d_kernel_perf}
  \caption{Performance of the Continuity and u-Momentum kernels as a
    function of SIMD vectorisation and problem size. Results are for
    v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.{\bf This plot
requires updating or dropping in favour of the clustered-bar form.}}
  \label{FIG_kernel_perf}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=130mm]{continuity_opt}
  \caption{Performance of the Continuity kernel with optimisations
    intended to aid SIMD vectorisation. Results are for v.16 of the
    Intel compiler on a Xeon E5-1620 v.2 CPU.}
  \label{FIG_cont_vec}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=120mm]{nemolite2d_cont_perf}
  \caption{Performance of the Continuity kernel (open circles -- right y-axis)
    and associated HWPC data (bars -- left y-axis) as a function of
    problem size. Results are for v.16 of the Intel compiler on a Xeon
    E5-1620 v.2 CPU.}
  \label{FIG_cont_perf}
\end{figure}
